### 2020-07-09 paper.tex
Finish the proof sketch and fix inverse error

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $u \sim q(\cdot; x)$ \Comment{Sample auxiliary variables}
    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
    %\State $\alpha \gets
        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
    % TODO or should we show the Radon-Nikodym derivative?
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)^{-1}$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}

@@

\section{Proof of detailed balance for involution}
Detailed balance is proven by applying a result from~\citet{tierney1998note} for general Metropolis-Hastings moves with deterministic proposals, and using the measure $h(z) := \pi(z) / (\pi(z) + \pi'(z))$ where $\pi'$ 
% See top of page 4, bullet point 2, of Tierney 1998 note

\section{Proof of stationarity for involution}
Stationarity derives from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\section{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various conditions

>>>>

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $u \sim q(\cdot; x)$ \Comment{Sample auxiliary variables}
    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
    %\State $\alpha \gets
        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
    % TODO or should we show the Radon-Nikodym derivative?
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}

@@

\section{Derivation of pushforward Radon-Nikodym derivative for special case}
TODO: Do some special case where it simplifies to a Jacobian determinant.
Maybe the case where the involution is composed of an involution on discrete part and a diffeomorphism on the continuous part.

\section{Proof of detailed balance for involution}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
$\pi'(A) := \pi(T^{-1}(A))$.
They let $\nu(A) := \pi(A) + \pi'(A)$.
They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
Then, they set:
\[
\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
\]
Our definition of $\alpha$ satisfies:
\[
\frac{\alpha(z \to T(z))}{\alpha(T(z) \to z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
Therefore, to use Tierney's result, it suffices to show that (when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$):
\[
\frac{h(T(z))}{h(z)}= \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
(note that $T$ is our $f$).
What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$, so the density of $\pi$ with respect to $\nu$ is:
\[
h(z) = \frac{\pi(z)}{\nu(z)} %= \frac{\pi(z)}{\pi(z) + \pi'(z)}
\]
and therefore:
\[
\frac{h(T(z))}{h(z)} = \frac{\pi(T(z))}{\pi(z)} \cdot \frac{\nu(z)}{\nu(T(z))}
\]
It suffices to show that:
\[
\frac{\nu(T(z))}{\nu(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
First, note that $\nu$ and $\nu \circ T^{-1}$ are the same measure:
\begin{align}
\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(T^{-1}(A))\\
(\nu \circ T^{-1})(A) &= \pi(T^{-1}(A)) + \pi(T^{-1}(T^{-1}(A))) = \pi(T^{-1}(A)) + \pi(A)
\end{align}
Therefore,
\[
\frac{d (\nu \circ T^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
\]
Expanding $d(\nu \circ T^{-1}) / d \nu$ using the chain rule:
\begin{align}
1 &= \frac{d (\nu \circ T^{-1})}{d \nu}(z)\\
&=   \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) \cdot
    \frac{d (\mu \circ T^{-1})}{d \mu}(z) \cdot
    \frac{d \mu}{d \nu}(z)\\
&= \nu(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{\nu(z)}
\end{align}
where in the second step we used that
\[
\frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
\]
To show this, use (i) $\nu \circ T^{-1} = \nu$, (ii) the definition of Radon-Nikodym derivative $d \nu / d \mu$ and (iii) and Theorem 3.6.1 in Bogachev with $g := d\nu / d\mu$:
\begin{align}
(\nu \circ T^{-1})(A) = \nu(A) &= \int_A \frac{d \nu}{d \mu}(z) \mu(dz) = \int_A \frac{d \nu}{d \mu} (T(z)) ( \mu \circ T^{-1} )(dz)
\end{aling}

\section{Proof of stationarity for involution}
Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\section{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various assumptions that allow us to e.g. rearrange the order of integration, etc.
